Computational methods in landscape ecology

### 2024-09-19 Sections/08-discussion.qmd
Remove trailing whitespaces
Update Discussion

Since emerging in the 1980s <!-- REF -->, the field of landscape ecology has constantly evolved [@Costanza2019DescribingAnalyzing; @Gustafson2019HowHas] and is now highly depending on computational methods.
Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
To maximize the benefits of "Open science", such as shareability, reproducibility, and transparency, the research community will gain the most if both existing and new software support script-based workflows and are made available under open-source licenses [@Powers2019OpenScience; @Vogt2019PatternsSoftware].
<!--JN: maybe add a sentence or two summarizing the previous sections (e.g., that this paper presents various computational methods used in landscape ecology)-->

<!-- JN: the next sentence seems to be broken -->
Furthermore, landscape ecology will benefit and adapt solutions from technological advancements in other fields.
This includes advancements in remote sensing technology, with increased data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems. 
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR[@Crowley2020RemoteSensings].
In combination with other increasingly available data sources, e.g, citizen science data, ecology is in the era of increasingly large data requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modeling, or prediction and simulations [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].

<!-- MH: Add refs to this -->
Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them.
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution.
However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions. 
Here we provided a review of the latest developments, nevertheless, a review of these methods can always be only a snapshot in time and hopefully new computational approaches will emerge in the future.

<!-- JN: additional possible topics to discuss here is that the computational methods are also related to the developments in computer science; thus, new frameworks, tools, good practices, etc. are constantly emerging, and not everything is about new methods, but also about new ways to implement them, new ways to share them, better ways to document them, more consistent syntaxes, etc. -->

>>>>

Since emerging in the 1980s [@With2019EssentialsLandscape], the field of landscape ecology has constantly evolved and is now highly depending on computational methods.
Thus, future progress within the research field will largely depend on available software and analysis tools [@Costanza2019DescribingAnalyzing].
Here, we present and summarize some recent developments of computational methods related to spatial patterns, connectivity, simulations and modelling, landscape genetics and spatial planning.
In this context, we highlight open-source software as a cornerstone of Open Science, offering key advantages like shareability, reproducibility, and transparency, which provide the greatest benefits to the research community [@Powers2019OpenScience; @Vogt2019PatternsSoftware].

Alongside methodological advancements specific to landscape ecology, the research field will also benefit from and integrate technological innovations from related disciplines.
This includes advancements in remote sensing technology, with increasing data availability and diversity from a range of sources, including large satellites, unmanned aerial vehicles, and ground-based sensor systems.
These systems provide both passive sensors such as multispectral, hyperspectral, and thermal sensors, and active optical sensors such as LiDAR and SAR[@Crowley2020RemoteSensings].
In combination with other increasingly available data sources, e.g, citizen science data, ecology is entering the era of big data, requiring methods that can handle heterogeneous data or high-throughput computing resources [@Farley2018SituatingEcology].
Due to its high flexibility and performance, artificial intelligence has become popular and deep learning and machine learning algorithms can be used for e.g., mapping, classification and feature extraction, modelling, or predicting [@Christin2019ApplicationsDeep; @Stupariu2022MachineLearning; @Pichler2023MachineLearning].
Quantum computing may additionally offer a potential pathway to surpass the limits of current computational technologies [@Woolnough2023QuantumComputing].
Furthermore, landscape ecology will benefit from a general paradigm shift related to data sharing, management, documentation [@Tenopir2020DataSharing].

Climate change as well as the biodiversity crisis are two intervened and complex issues that need integrative, multi-disciplinary, and scale-dependent solutions to face them [@Pettorelli2021TimeIntegrate; @Portner2023OvercomingCoupled].
Landscape ecology is well equipped to provide answers because it connects several research fields at multiple scales, such as social sciences, geography, and ecology and evolution.
However, up-to-date and constantly evolving computational methods are required to meet the increasing complexity of research questions.
Here we provided a review of the latest developments, nevertheless, a review of these methods can always be only a snapshot in time and hopefully new computational approaches will emerge in the future.
